Short-term synaptic plasticity and threshold adaptation act in synergy to give rise to deviance detection in a spiking model of neuronal culture

### 2022-12-08 paper-1/manuscript.tex
Rev sec Results.DD

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{\textbf{a} Illustration of the oddball paradigm used. Each square represents a stimulus, with the target (A) in dark blue, and other stimuli in lighter colors. The temporal extent of stimulation is exaggerated; in reality, stimulation was delivered in a single time step, and the response allowed to develop freely for 500 ms. \textbf{b} Response of a neuronal cell culture in vitro, adapted from \cite{Kubota2021-dx}. \textbf{c} Model response, showing average network-wide spike counts after stimulation with target (A) in each sequence. \textbf{d} Mismatch, SSA and deviance detection indices across networks and stimulus pairs (n = 120). In this and later figures, notches indicate the confidence interval for the median, calculated by bootstrap with 10000 iterations. All indices were positive (mismatch: t = 11, p = 6.7e-21, SSA: t = 7.8, p = 1.05e-12, deviance detection: t = 6.5, p = 1.05e-09)}
    \label{fig2}
\end{figure*}

\textbf{Fig. 2 a} Illustration of the oddball paradigm used. Each square represents a stimulus, with the target (A) in dark blue, and other stimuli in lighter colors. The temporal extent of stimulation is exaggerated; in reality, stimulation was delivered in a single time step, and the response allowed to develop freely for 500 ms. \textbf{b} Response of a neuronal cell culture in vitro, adapted from \cite{Kubota2021-dx}. \textbf{c} Model response, showing average network-wide spike counts after stimulation with target (A) in each sequence. \textbf{d} Mismatch, SSA and deviance detection indices across networks and stimulus pairs (n = 120). In this and later figures, notches indicate the confidence interval for the median, calculated by bootstrap with 10000 iterations. All indices were positive (mismatch: t = 11, p = 6.7e-21, SSA: t = 7.8, p = 1.05e-12, deviance detection: t = 6.5, p = 1.05e-09)

We define deviance detection as a network's ability to pick out unexpected inputs from an otherwise homogeneous or otherwise unsurprising stream of inputs. Following related research \citep{TODO}, we were careful to exclude effects due to stimulus identity (i.e., avoiding direct comparisons between different stimuli) or due to adaptation to frequent stimulus presentation (i.e., avoiding direct comparisons of A in the dev sequence to A in the std sequence). In other words, we used the msc sequence as a neutral baseline where the target stimulus was presented no more often than in the dev sequence, but no expectation of a competing stimulus B could be formed. To quantify deviance detection, we compared the mean number of spikes fired across the network in response to target stimulus A in the dev and msc sequences, defining a deviance detection index (DDI) as
\begin{equation}
    DDI = \frac{R^A_{dev} - R^A_{msc}}{R^A_{dev} + R^A_{msc}}
\end{equation}
We define stimulus-specific adaptation (SSA, msc - std) and mismatch (dev - std) indices analogously, capturing respectively the effect of adaptation, and of both adaptation and deviance detection together.

Our first aim was to replicate deviance detection properties reported in dissociated neuronal cell culture by \cite{Kubota2021-dx}, reproduced in Figure \ref{fig2}b. With minimal tuning of global parameters, we found that most networks (92 out of 120 target stimuli) reliably produced not only mismatch responses based on adaptation alone, but also showed robust deviance detection, as summarized in Figure \ref{fig2}d.

The model responses developed on a compressed time scale compared to neuronal culture, which we tentatively attribute to unfaithful aspects of the model, including a lack of conduction delay or spontaneous activity, a much smaller number of participating neurons, and structural differences. However, we could readily identify target stimuli whose response profiles looked otherwise qualitatively similar to the cell culture results.

Having confirmed the presence of deviance detection in our model, we then turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model -- synaptic short-term depression (STD) and threshold adaptation (TA) -- enabled the networks to learn relevant regularities. As shown in Figure \ref{fig3}, networks with neither plasticity mechanism did not show any difference in response, indicating that stimulation did not have any long-lasting effects in membrane voltage or synaptic conductance. Networks with only STD likewise did not show any deviance detection, though they did exhibit robust stimulus-specific adaptation effects, leading to a smaller target response in std than in either dev or msc sequences. Conversely, with only TA, networks produced both adaptation and deviance detection effects.

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{Deviance detection and SSA indices under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. \textbf{b, c} Deviance detection and SSA indices across networks and stimulus pairs. Significance was assessed by 1-sample t-tests. TODO p-values: text or table?}
    \label{fig3}
\end{figure*}

\textbf{Fig. 3} Deviance detection and SSA indices under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. \textbf{b, c} Deviance detection and SSA indices across networks and stimulus pairs. Significance was assessed by 1-sample t-tests. TODO p-values: text or table?

Finally, with both STD and TA applied, deviance detection was also reliable, as shown before; however, we were surprised to find that the effect was significantly stronger than that of either plasticity mechanism alone (STD: t = 6.7, p = 4.53e-10; TA: t = 3.2, p = 0.00103), and greater too than the sum of the DDI values across both STD only and TA only models (t = 3.6, p = 0.000199). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

In the following, we will use the sample network and target stimulus shown in Figure \ref{fig3}a to first build an understanding of how TA leads to deviance detection when acting alone. Then, we will build on this understanding to elucidate how the addition of STD, which is ineffective on its own, can lead to an increased deviant response. The particular network and target stimulus site used in the following individual analyses was selected for the response size to target and non-target stimuli (A, B), both of which were average, as well as its deviance detection performance, which likewise reflected the tendencies observed across all networks.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig2.pdf}
    \caption{}
    \label{fig2}
\end{figure*}
\textbf{Fig. 2} Deviance detection in vitro and in silico. \textbf{a} Response of a 2 \times 4~mm neuronal cell culture on a multi-electrode array, adapted from \cite{Kubota2021-dx}. Curves represent the average number of spikes per electrode in each 10~ms bin in response to target stimulation in the indicated sequence. Unlike in our study, oddball sequences here were split 10\%/90\%, and 10 equally frequent stimuli were used for the msc sequence. \textbf{b} Sample model response, showing average network-wide spike counts after stimulation with target (A) in the indicated sequence (same color code as panel \textbf{a}). Both in vitro and in the model, the dev response is very large, followed by an intermediate msc response and a low std response. \textbf{c} Mismatch, SSA and deviance detection indices (see Equation \ref{eqn-ddi}) across networks and stimuli (n = 120). In this and later boxplots, the median is indicated with an orange line, notches indicate the confidence interval for the median calculated by bootstrap with 10000 iterations, boxes indicate the inter-quartile range (IQR), whiskers extend to 1.5 times the IQR or to the data extrema, whichever is less, and fliers indicate individual data points beyond the whisker end points

We define deviance detection as a network's ability to pick out unexpected inputs from an otherwise homogeneous or otherwise unsurprising stream of inputs. Following related research \citep{Kubota2021-dx,Harms2014-ah,Jacobsen2001-sc}, we were careful to exclude effects due to stimulus identity (i.e., avoiding direct comparisons between different stimuli) or due to adaptation to frequent stimulus presentation (i.e., avoiding direct comparisons of A in the dev sequence to A in the std sequence). In other words, we used the msc sequence as a neutral baseline where the target stimulus was presented no more often than in the dev sequence, but no expectation of a competing stimulus B could be formed. To quantify deviance detection, we compared the mean number of spikes $R^A$ fired across the network in response to target stimulus A in the dev and msc sequences, defining a deviance detection index (DDI) as
\begin{equation}
    DDI = \frac{R^A_{dev} - R^A_{msc}}{R^A_{dev} + R^A_{msc}} \label{eqn-ddi}
\end{equation}
We define stimulus-specific adaptation (SSA, $R^A_{msc} - R^A_{std}$) and mismatch ($R^A_{dev} - R^A_{std}$) indices analogously, capturing respectively the effect of adaptation, and of both adaptation and deviance detection together.

Our first aim was to replicate deviance detection properties reported in dissociated neuronal cell culture by \cite{Kubota2021-dx}, reproduced in Figure \ref{fig2}a. In their case, cell cultures -- like our model, randomly connected networks with two-dimensional extent -- were stimulated at 10 different sites in msc sequences, and at two sites with 10\%/90\% probability in dev and std sequences. Deviance detection was assessed analogously to our definition of DDI and reached an average level of around 0.3 across two cultures.

In our model, we could readily identify networks and stimuli whose responses developed in a qualitatively similar fashion, with high activity in response to target in the dev sequence, low activity in the std sequence, and intermediate activity in the msc sequence, as illustrated with an example in Figure \ref{fig2}b. The model responses developed on a compressed time scale compared to neuronal culture, which we tentatively attribute to unfaithful aspects of the model, including a lack of conduction delay or spontaneous activity, a much smaller number of participating neurons, and structural differences.

Across networks and stimuli (Figure \ref{fig2}c), we found that our model reliably produced mismatch responses (t = 11.3, p = 6.7e-21; one-sided t-test for mismatch index > 0, n = 120) while both SSA index (t = 7.84, p = 1.05e-12) and DDI (t = 6.49, p = 1.05e-09) were also positive. The average DDI value reached around 0.1, which was less than the values reported by \cite{Kubota2021-dx}, but note that our target stimuli were presented more frequently (20\%) than theirs (10\%) in the dev and msc sequences.

\begin{figure*}%
    \centering
    \includegraphics{fig3.pdf}
    \caption{}
    \label{fig3}
\end{figure*}
\textbf{Fig. 3} Deviance detection under model ablation. \textbf{a} Sample network responses to target stimulation, with the deviance detection index in each instance noted in the plot titles. The color code is the same as in Figure \ref{fig2}. \textbf{b} SSA indices across all networks and stimuli. Asterisks indicate data greater than 0 at a significance level of p<0.05, see main text for detailed statistics. \textbf{c} Deviance detection indices across all networks and stimuli.

Having confirmed the presence of deviance detection in our model, we then turned to an ablation approach to try to identify how the two short-term plasticity mechanisms in the model, STD and TA, enabled the networks to learn relevant regularities. Figure \ref{fig3}a shows the responses to target stimulation of a sample network under ablation. With no plasticity, the responses are indistinguishable between sequences, yielding a DDI of 0, and indicating that stimulation did not have any long-lasting effects in membrane voltage or synaptic conductance. With STD only, the response in the dev sequence developed earlier and very slightly larger, yielding a DDI of 0.03. With TA only, the response became noticeably smaller, but the sequences became readily distinguishable, with high dev, intermediate msc, and low std responses, yielding a DDI of 0.19. Finally, in the full model, the dev response stood out more clearly, leaving much smaller msc and std responses and yielding a DDI of 0.4.

The SSA index (Figure \ref{fig3}b) was exactly 0 as expected in the ablation with no plasticity, and positive with any plasticity applied (STD only: t = 3.17, p = 0.000983, median 0.0039; TA only: t = 6.74, p = 3e-10, median 0.097; STD+TA: t = 7.84, p = 1.05e-12, median 0.135; one-sided t-tests, n = 120). Since the SSA index measures the degree of adaptation to frequent stimulus presentation, this indicates that both plasticity mechanisms were capable of imbuing the networks with some level of history dependence.

Finally, across networks and stimuli, the DDI (Figure \ref{fig3}c) was exactly 0 with no plasticity, not significantly different from 0 with STD only (t = -1.89, p = 0.0609, median -2.1e-04; two-sided t-test, n = 120), but clearly positive for TA only (t = 5.52, p = 1.01e-07, median 0.077; one-sided t-test, n = 120) and the full model (t = 6.49, p = 1.05e-09, median 0.114). To our surprise, the DDI of the full model was significantly greater than that of either plasticity mechanism alone (STD: t = 6.7, p = 4.53e-10; TA: t = 3.2, p = 0.00103), and greater too than the sum of the DDI values across both STD only and TA only models (t = 3.6, p = 0.000199). This suggests that TA and STD act not as independent mechanisms in this paradigm, but rather interact constructively to enhance the network's ability to encode input regularities.

In the following, we will use a sample network and target stimulus to first build an understanding of how TA leads to deviance detection when acting alone, i.e. in the ablated TA only model. Then, we will build on this understanding to elucidate how the addition of STD, which was shown above to be ineffective for deviance detection on its own, can lead to an increased deviant response in the full model.
